---
output: pdf_document
---
## Reproducible Research: Peer Assessment 2

### Analysis Severe Weather Impact on Public Health and Economy

###Synonpsis

Based on the storm data collected from the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011, this report analyzes the impact of different weather events on public health and economy. It is to answer to two questions:  
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  
2. Across the United States, which types of events have the greatest economic consequences?


### Data Processing
```{r libfrary}
library(R.utils)
library(ggplot2)
library(plyr)
require(gridExtra)
```
#### Loading the data
```{r LoadingData}
stormData <- read.csv("stormData.csv")
head(stormData, n=3)
```
Preparing fatalities and injuries data
```{r pupHealth}
sortAid <- function(fieldName, top = 15, dataset = stormData) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

fatalities <- sortAid("FATALITIES", dataset = stormData)
injuries <- sortAid("INJURIES", dataset = stormData)
```

Preparing property and crop damage data
```{r econImpct}
convertAid <- function(dataset = stormData, fieldName, newFieldName) {
    totalLen <- dim(dataset)[2]
    index <- which(colnames(dataset) == fieldName)
    dataset[, index] <- as.character(dataset[, index])
    logic <- !is.na(toupper(dataset[, index]))
    dataset[logic & toupper(dataset[, index]) == "B", index] <- "9"
    dataset[logic & toupper(dataset[, index]) == "M", index] <- "6"
    dataset[logic & toupper(dataset[, index]) == "K", index] <- "3"
    dataset[logic & toupper(dataset[, index]) == "H", index] <- "2"
    dataset[logic & toupper(dataset[, index]) == "", index] <- "0"
    dataset[, index] <- as.numeric(dataset[, index])
    dataset[is.na(dataset[, index]), index] <- 0
    dataset <- cbind(dataset, dataset[, index - 1] * 10^dataset[, index])
    names(dataset)[totalLen + 1] <- newFieldName
    return(dataset)
}

stormData <- convertAid(stormData, "PROPDMGEXP", "propertyDamage")
stormData <- convertAid(stormData, "CROPDMGEXP", "cropDamage")
names(stormData)
property <- sortAid("propertyDamage", dataset = stormData)
crop <- sortAid("cropDamage", dataset = stormData)
property$propertyDamage <- property$propertyDamage/10^6
crop$cropDamage <- crop$cropDamage/10^6
```


### Results

#### Aross the United States, Which types of events are nost harmful with respect to population health?
```{r plots}
fatalitiesPlot <- qplot(EVTYPE, data = fatalities, weight = FATALITIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Fatalities") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Fatalities by Severe Weather")
injuriesPlot <- qplot(EVTYPE, data = injuries, weight = INJURIES, geom = "bar", binwidth = 1) + 
    scale_y_continuous("Number of Injuries") + 
    theme(axis.text.x = element_text(angle = 45, 
    hjust = 1)) + xlab("Severe Weather Type") + 
    ggtitle("Total Injuries by Severe Weather")
grid.arrange(fatalitiesPlot, injuriesPlot, ncol = 2)
```

The most harmful weather event to population health is Tornado.It is cause for both the highest fatalities and the highest injuries across United States.

#### Across the United States, which types of events have the greatest economic consequences?
```{r plots2}
propertyPlot <- qplot(EVTYPE, data = property, weight = propertyDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Property Damage (million $)")+ 
    xlab("Severe Weather Type") + ggtitle("Total Property Damage by\n Severe Weather Events ")

cropPlot<- qplot(EVTYPE, data = crop, weight = cropDamage, geom = "bar", binwidth = 1) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + scale_y_continuous("Crop Damage (million $)") + 
    xlab("Severe Weather Type") + ggtitle("Total Crop Damage by \nSevere Weather Events")
grid.arrange(propertyPlot, cropPlot, ncol = 2)
```

The weather events have the greatest economic consequences are: flood, drought, Tornado and Typhoon.

Across the United States, flood, tornado and typhoon have caused the greatest damage to properties. Drought and flood come as the causes for the greatest damage to crops.